Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-30 doc/paper.Rmd
Edits to results, survival, growth and flowering (exploratory data analysis section)

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(knitr)
library(here)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot,
    eda_plot,
    s_spei_plot,
    s_spei_diff_plot,
    g_spei_plot,
    g_spei_diff_plot,
    f_spei_plot,
    f_spei_diff_plot,
    size_plot,
    f_cf,
    f_1ha
  ))
})
```

@@

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN. Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap? -->

<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

@@

<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot,
    eda_plot,
    s_spei_plot,
    s_spei_diff_plot,
    g_spei_plot,
    g_spei_diff_plot,
    f_spei_plot,
    f_spei_diff_plot,
    size_plot,
    f_cf,
    f_1ha,
    model_data,
    xa_spei
  ))
})
```

@@

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

```{r include=FALSE}
summary_stats <-
  model_data %>%
  group_by(habitat) %>%
  summarize(across(c(ht, shts, size, flwr, surv),
                   .fns = list(mean = mean, sd = sd),
                   na.rm = TRUE)) %>%
  right_join(model_data %>% count(habitat)) %>%
  mutate(across(ends_with("_mean") | ends_with("_sd"),
                                ~round(.x, 2)))
summary_stats <- 
  full_join(
  summary_stats,
  model_data %>%
    count(habitat, plot) %>%
    group_by(habitat) %>%
    summarize(n_mean = mean(n), n_sd = sd(n)) %>%
    mutate(across(where(is.numeric), round))
)
summary_stats

by_year <-
  model_data %>% 
  group_by(year, habitat) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
by_year

xa_spei %>% ungroup() %>%
  summarize(across(everything(), mean)) %>%
  filter(year(yearmonth)==2005)

Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

@@
